---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# kread

<!-- badges: start -->
[![Release](https://gitlab.com/OrlandoKevin/kread/-/badges/release.svg)](https://gitlab.com/OrlandoKevin/kread.git)
[![License: AGPL v3](https://img.shields.io/badge/License-AGPLv3-blue.svg)](https://www.gnu.org/licenses/agpl-3.0)
[![Experimental](https://img.shields.io/badge/status-experimental-orange.svg)](https://gitlab.com/OrlandoKevin/kread)
[![pipeline status](https://gitlab.com/OrlandoKevin/kread/badges/main/pipeline.svg)](https://gitlab.com/OrlandoKevin/kread/commits/main)
[![Dev coverage status](https://gitlab.com/OrlandoKevin/kread/badges/main/coverage.svg)](https://orlandokevin.gitlab.io/kread/coverage.html)
<!-- badges: end -->

The `kread` package provides a unified interface for reading various file formats into R. Instead of remembering which function to use for each file type, `kread` automatically detects the file format based on its extension and applies the appropriate reading function.

## Features

- **Automatic format detection**: Just use `read()` and let the package figure out the format
- **Wide format support**: CSV, TSV, Excel, ODS, NetCDF, Shapefiles, RDS, YAML, INI
- **Sensible defaults**: Suppresses verbose messages by default
- **Multi-sheet Excel support**: Read one or all sheets with a single function call
- **Sheet-specific arguments**: Apply different parameters to different Excel sheets
- **Consistent API**: Same interface across all file types

## Installation

You can install the development version of `kread` from GitLab:

```{r eval=FALSE}
# install.packages("remotes")
remotes::install_gitlab("OrlandoKevin/kread")
```

Or using git URL:

```{r eval=FALSE}
remotes::install_git(
  "https://gitlab.com/OrlandoKevin/kread.git", 
  dependencies = TRUE
)
```

## Supported File Formats

| Format | Extensions | Function Used |
|--------|-----------|---------------|
| Delimited files | `.csv`, `.tsv`, `.dat` | `readr::read_delim` |
| Excel | `.xls`, `.xlsx` | `readxl::read_excel` |
| OpenDocument | `.ods` | `readODS::read_ods` |
| Shapefiles | `.shp` | `sf::read_sf` |
| NetCDF | `.nc` | `stars::read_mdim` |
| R objects | `.rds`, `.RDS` | `readRDS` |
| YAML | `.yml`, `.yaml` | `yaml::read_yaml` |
| INI | `.ini` | `ini::read.ini` |

## Quick Start

```{r eval=FALSE}
library(kread)

# Read any file format with automatic detection
data_csv <- read("data.csv")
data_excel <- read("report.xlsx")
data_spatial <- read("boundaries.shp")

# Read with additional arguments
data <- read("data.csv", delim = ";", skip = 2)

# Read Excel with multiple sheets
all_sheets <- read_excel("workbook.xlsx")
specific_sheets <- read_excel("workbook.xlsx", sheets = c("Data", "Summary"))
```

## Main Functions

### `read()` - Universal File Reader

Automatically detects file type and uses the appropriate reading function:

```{r eval=FALSE}
# Automatic format detection
data <- read("myfile.csv")
data <- read("myfile.xlsx")
data <- read("myfile.shp")

# Pass arguments to underlying function
data <- read("myfile.csv", delim = ",", col_names = TRUE)

# Override automatic detection
data <- read("myfile.txt", FUN = readr::read_csv)
```

### `read_table()` - Delimited Files

Specialized function for reading delimited files (CSV, TSV, etc.) with sensible defaults:

```{r eval=FALSE}
# Read CSV quietly (no column type messages)
data <- read_table("data.csv")

# Specify delimiter
data <- read_table("data.tsv", delim = "\t")

# Show column types if needed
data <- read_table("data.csv", show_col_types = TRUE)
```

### `read_excel()` - Excel Files

Enhanced Excel reading with multi-sheet support:

```{r eval=FALSE}
# Read all sheets (returns a list)
all_data <- read_excel("workbook.xlsx")

# Read specific sheets
data <- read_excel("workbook.xlsx", sheets = c("Sheet1", "Sheet2"))

# Read single sheet (returns data frame, not list)
data <- read_excel("workbook.xlsx", sheets = "Data")

# Apply different arguments to different sheets
data <- read_excel(
  "workbook.xlsx",
  sheets = c("Sheet1", "Sheet2"),
  Sheet1 = list(skip = 2),
  Sheet2 = list(skip = 0)
)
```

## Examples

### Reading Multiple Files

```{r eval=FALSE}
# Read all CSV files in a directory
files <- list.files("data/", pattern = "\\.csv$", full.names = TRUE)
data_list <- lapply(files, read)
combined <- do.call(rbind, data_list)
```

### Working with Geospatial Data

```{r eval=FALSE}
# Read shapefile
boundaries <- read("admin_boundaries.shp")

# Read NetCDF climate data
climate <- read("temperature.nc")

# Works seamlessly with sf and stars packages
library(sf)
plot(boundaries)
```

### Configuration Files

```{r eval=FALSE}
# Read YAML config
config <- read("config.yml")

# Read INI file
database_config <- read("database.ini")
```

## Comparison with Traditional Approach

**Before (Traditional):**
```{r eval=FALSE}
csv_data <- read.csv("data.csv")
excel_data <- readxl::read_excel("data.xlsx")
yaml_data <- yaml::read_yaml("config.yml")
spatial_data <- sf::read_sf("boundaries.shp")
```

**After (with kread):**
```{r eval=FALSE}
csv_data <- read("data.csv")
excel_data <- read("data.xlsx")
yaml_data <- read("config.yml")
spatial_data <- read("boundaries.shp")
```

## Documentation

For more detailed information:

- Run `?read`, `?read_table`, or `?read_excel` in R
- Browse the [vignette](vignettes/kread-intro.Rmd) for comprehensive examples
- Visit the [package website](https://orlandokevin.gitlab.io/kread/) (when available)

## Contributing

Contributions are welcome! Please feel free to submit issues or merge requests on [GitLab](https://gitlab.com/OrlandoKevin/kread).

## License

This package is licensed under AGPL-3.0 or later.
