---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# kread

<!-- badges: start -->
[![Release](https://gitlab.com/kbosirany/kread/-/badges/release.svg)](https://gitlab.com/kbosirany/kread.git)
[![License: AGPL v3](https://img.shields.io/badge/License-AGPLv3-blue.svg)](https://www.gnu.org/licenses/agpl-3.0)
[![Experimental](https://img.shields.io/badge/status-experimental-orange.svg)](https://gitlab.com/kbosirany/kread)
[![pipeline status](https://gitlab.com/kbosirany/kread/badges/main/pipeline.svg)](https://gitlab.com/kbosirany/kread/commits/main)
[![Dev coverage status](https://gitlab.com/kbosirany/kread/badges/main/coverage.svg)](https://kbosirany.gitlab.io/kread/coverage.html)
<!-- badges: end -->

The `kread` package provides a unified interface for reading various file
formats into R. Instead of remembering which function to use for each file type,
`kread` automatically detects the file format based on its extension and applies
the appropriate reading function.

## Features

- **Automatic format detection**: Just use `read()` and let the package figure
out the format
- **Wide format support**: CSV, TSV, Excel, ODS, NetCDF, Shapefiles, RDS, YAML,
INI
- **Sensible defaults**: Suppresses verbose messages by default
- **Multi-sheet Excel support**: Read one or all sheets with a single function
call
- **Sheet-specific arguments**: Apply different parameters to different Excel
sheets
- **Consistent API**: Same interface across all file types

## Installation

You can install the development version of `kread` from GitLab:

```{r eval=FALSE}
# install.packages("remotes")
remotes::install_gitlab("kbosirany/kread")
```

Or using git URL:

```{r eval=FALSE}
remotes::install_git(
  "https://gitlab.com/kbosirany/kread.git", 
  dependencies = TRUE
)
```

## Supported File Formats

| Format | Extensions | Function Used |
|--------|-----------|---------------|
| Delimited files | `.csv`, `.tsv`, `.dat` | `readr::read_delim` |
| Excel | `.xls`, `.xlsx` | `readxl::read_excel` |
| OpenDocument | `.ods` | `readODS::read_ods` |
| Shapefiles | `.shp` | `sf::read_sf` |
| NetCDF | `.nc` | `stars::read_mdim` |
| R objects | `.rds`, `.RDS` | `readRDS` |
| YAML | `.yml`, `.yaml` | `yaml::read_yaml` |
| INI | `.ini` | `ini::read.ini` |

## Comparison with Traditional Approach

**Before (Traditional):**
```{r eval=FALSE}
csv_data <- read.csv("data.csv")
excel_data <- readxl::read_excel("data.xlsx")
yaml_data <- yaml::read_yaml("config.yml")
spatial_data <- sf::read_sf("boundaries.shp")
```

**After (with kread):**
```{r eval=FALSE}
csv_data <- read("data.csv")
excel_data <- read("data.xlsx")
yaml_data <- read("config.yml")
spatial_data <- read("boundaries.shp")
```

## Documentation

For more detailed information:

- Run `?read`, `?read_table`, or `?read_excel` in R
- Browse the [vignette](vignettes/kread-intro.Rmd) for comprehensive examples
- Visit the [package website](https://kbosirany.gitlab.io/kread/) (when
available)

## Contributing

Contributions are welcome! Please feel free to submit issues or merge requests
on [GitLab](https://gitlab.com/kbosirany/kread).

## License

This package is licensed under AGPL-3.0 or later.
