---
title: "Introduction to kread"
author: "Kevin Bosirany Orlando"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to kread}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(kread)

# Create temporary files for examples
temp_dir <- tempdir()

# Create a sample CSV file
csv_file <- file.path(temp_dir, "sample.csv")
write.csv(data.frame(
  id = 1:5,
  name = c("Alice", "Bob", "Charlie", "David", "Eve"),
  score = c(85, 92, 78, 95, 88)
), csv_file, row.names = FALSE)

# Create a sample TSV file
tsv_file <- file.path(temp_dir, "sample.tsv")
write.table(data.frame(
  product = c("A", "B", "C"),
  price = c(10.5, 20.3, 15.7),
  quantity = c(100, 50, 75)
), tsv_file, sep = "\t", row.names = FALSE)

# Create a sample RDS file
rds_file <- file.path(temp_dir, "sample.rds")
saveRDS(mtcars[1:10, ], rds_file)
```

# Introduction

The `kread` package provides a unified interface for reading various file formats into R. Instead of remembering which function to use for each file type, `kread` automatically detects the file format based on its extension and applies the appropriate reading function.

## Installation

You can install the development version of kread from GitLab:

```{r installation, eval = FALSE}
# install.packages("remotes")
remotes::install_gitlab("OrlandoKevin/kread")
```

## Supported File Formats

`kread` supports a wide range of file formats:

- **Delimited files**: CSV, TSV, DAT
- **Spreadsheets**: Excel (.xls, .xlsx), OpenDocument Spreadsheet (.ods)
- **Geospatial**: Shapefiles (.shp), NetCDF (.nc)
- **R objects**: RDS files
- **Configuration**: YAML (.yml, .yaml), INI (.ini)

## Basic Usage

### The `read()` Function

The main function is `read()`, which automatically detects the file type and uses the appropriate reader:

```{r basic-read}
# Read a CSV file
data_csv <- read(csv_file, delim = ",")
head(data_csv)

# Read an RDS file
r_object <- read(rds_file)
head(r_object)
```

The function automatically detects the file format based on the extension and uses the appropriate reader. For demonstration purposes with files that aren't available in this vignette:

```{r basic-read-demo, eval = FALSE}
# Read an Excel file
data_excel <- read("data/myfile.xlsx")

# Read a shapefile
spatial_data <- read("data/boundaries.shp")

# Read a YAML configuration
config <- read("config/settings.yml")
```

### Passing Additional Arguments

You can pass additional arguments to the underlying reading functions:

```{r read-with-args}
# Read with specific delimiter
data <- read(csv_file, delim = ",", show_col_types = FALSE)

# Read specific columns (using readr syntax)
data_subset <- read(csv_file, delim = ",", col_select = c(id, name))
head(data_subset)
```

For files not available in this vignette:

```{r read-with-args-demo, eval = FALSE}
# Read CSV with semicolon delimiter
data <- read("data/european.csv", delim = ";")

# Skip first few rows in Excel
data <- read("data/report.xlsx", skip = 3)
```

## Specialized Reading Functions

### Reading Delimited Files with `read_table()`

The `read_table()` function is specifically designed for delimited files and suppresses column type messages by default:

```{r read-table}
# Read a CSV file quietly (no column type messages)
data <- read_table(csv_file, delim = ",")
head(data)

# Read a TSV file with delimiter specification
data_tsv <- read_table(tsv_file, delim = "\t")
head(data_tsv)

# Show column types if needed
data_verbose <- read_table(csv_file, delim = ",", show_col_types = TRUE)
```

You can also specify custom column names:

```{r read-table-names, eval = FALSE}
# Specify custom column names
data_named <- read_table("data/myfile.csv", 
                         delim = ",",
                         col_names = c("ID", "Name", "Score"))
```

### Reading Excel Files with `read_excel()`

The `read_excel()` function provides enhanced functionality for working with Excel files, including multi-sheet support.

We'll demonstrate with the built-in example file from the `readxl` package:

```{r read-excel}
# Use readxl's example Excel file
example_file <- readxl::readxl_example("datasets.xlsx")

# See what sheets are available
readxl::excel_sheets(example_file)

# Read all sheets (returns a list)
all_sheets <- read_excel(example_file)
length(all_sheets)
names(all_sheets)

# Read specific sheets
selected_sheets <- read_excel(example_file, 
                              sheets = c("mtcars", "chickwts"))
lapply(selected_sheets, head, n = 3)

# Read a single sheet (returns a data frame, not a list)
single_sheet <- read_excel(example_file, sheets = "mtcars")
head(single_sheet)
```

### Sheet-Specific Arguments

You can apply different arguments to different sheets:

```{r sheet-specific}
# Apply different skip values to different sheets
data <- read_excel(
  example_file,
  sheets = c("mtcars", "chickwts"),
  mtcars = list(n_max = 5),      # Only read first 5 rows of mtcars
  chickwts = list(n_max = 3)     # Only read first 3 rows of chickwts
)

lapply(data, nrow)
lapply(data, head, n = 2)
```

For demonstration with other files:

```{r sheet-specific-demo, eval = FALSE}
# Mix sheet-specific and common arguments
data <- read_excel(
  "data/workbook.xlsx",
  sheets = c("Q1", "Q2", "Q3"),
  Q1 = list(skip = 5),            # Only Q1 skips 5 rows
  na = c("", "NA", "N/A")         # Applied to all sheets
)
```

## Advanced Usage

### Custom Reading Functions

If you need more control, you can specify a custom reading function:

```{r custom-function}
# Use a specific readr function
data <- read(csv_file, FUN = readr::read_csv)
head(data)

# Use a custom function
my_reader <- function(path) {
  readr::read_csv(path, col_types = readr::cols(.default = "c"), 
                  show_col_types = FALSE)
}
data <- read(csv_file, FUN = my_reader)
head(data)
```

### Getting the Reading Function

You can retrieve the function that would be used without actually reading the file:

```{r get-function}
# Get the appropriate reading function for different file types
get_read_fun(csv_file)
get_read_fun(rds_file)
get_read_fun("example.xlsx")
get_read_fun("config.yml")

# Use it later
read_fn <- get_read_fun(csv_file)
data <- read_fn(csv_file, delim = ",")
head(data)
```

## Practical Examples

### Example 1: Reading Multiple CSV Files

```{r example-multiple}
# Create multiple CSV files for demonstration
csv_files <- c(
  file.path(temp_dir, "data1.csv"),
  file.path(temp_dir, "data2.csv")
)

write.csv(data.frame(x = 1:3, y = letters[1:3]), csv_files[1], row.names = FALSE)
write.csv(data.frame(x = 4:6, y = letters[4:6]), csv_files[2], row.names = FALSE)

# Read all files into a list
all_data <- lapply(csv_files, read, delim = ",")

# Or with explicit function
all_data <- lapply(csv_files, read_table, delim = ",")

# Combine into one data frame
combined_data <- do.call(rbind, all_data)
combined_data
```

### Example 2: Working with Excel Workbooks

```{r example-excel}
# Read an Excel workbook with multiple sheets
workbook <- read_excel(example_file)

# Access individual sheets (returns unnamed list)
sheet1 <- workbook[[1]]
sheet2 <- workbook[[2]]

head(sheet1, 3)
head(sheet2, 3)

# Or read specific sheets by name
financial_data <- read_excel(
  example_file,
  sheets = c("mtcars", "chickwts")
)

# See the structure
str(financial_data, max.level = 1)
```

### Example 3: Reading Geospatial Data

Geospatial data reading (for demonstration - requires actual spatial files):

```{r example-spatial, eval = FALSE}
# Read a shapefile
boundaries <- read("data/admin_boundaries.shp")

# Read NetCDF climate data
climate_data <- read("data/temperature.nc")

# Both return spatial objects ready for analysis
library(sf)
plot(boundaries)
```

### Example 4: Configuration Files

Configuration file reading (for demonstration - requires actual config files):

```{r example-config, eval = FALSE}
# Read YAML configuration
config <- read("config/settings.yml")

# Read INI file
ini_config <- read("config/database.ini")

# Access configuration values
db_host <- ini_config$database$host
```

## Error Handling

`kread` will throw an error if the file extension is not supported:

```{r error-handling}
# This will error for unsupported formats
tryCatch(
  get_read_fun("document.pdf"),
  error = function(e) {
    message("Caught error: ", e$message)
  }
)

# Another example
tryCatch(
  get_read_fun("data.json"),
  error = function(e) {
    message("Caught error: ", e$message)
  }
)
```

## Comparison with Base R

### Traditional Approach
```{r traditional, eval = FALSE}
# You need to remember which function for which format
csv_data <- read.csv("data.csv")
excel_data <- readxl::read_excel("data.xlsx")
yaml_data <- yaml::read_yaml("config.yml")
rds_data <- readRDS("model.rds")
```

### With kread
```{r kread-approach}
# One function for all formats
csv_data <- read(csv_file, delim = ",")
excel_data <- read(example_file)
rds_data <- read(rds_file)

# All work the same way!
head(csv_data, 2)
head(rds_data, 2)
```

## Best Practices

1. **Use `read()` for quick exploration** - Let the automatic detection handle format selection
2. **Use specialized functions for production code** - `read_table()` and `read_excel()` offer more control
3. **Always handle errors** - Wrap file reading in `tryCatch()` for robust code
4. **Specify delimiters explicitly** - Even though `read_table()` can guess, being explicit prevents issues
5. **Document your file formats** - Keep track of what formats your workflow uses

## Performance Considerations

- For large CSV files, consider using `data.table::fread()` or `vroom::vroom()` directly
- For very large Excel files, consider converting to CSV first
- NetCDF and spatial files are loaded lazily by their respective packages

## Getting Help

For more information on specific functions:

```{r help, eval = FALSE}
?read
?read_table
?read_excel
?get_read_fun
```

## Summary

The `kread` package simplifies file reading in R by:

- Providing automatic format detection
- Offering a unified interface across file types
- Including sensible defaults (like suppressing column type messages)
- Supporting advanced features like multi-sheet Excel reading
- Maintaining compatibility with underlying reading functions

Whether you're working with simple CSV files or complex multi-sheet Excel workbooks, `kread` makes data import straightforward and consistent.

```{r cleanup, include = FALSE}
# Clean up temporary files
unlink(c(csv_file, tsv_file, rds_file, csv_files))
```
